---
title: "MATH390: Homework 3 - 2020 Week 1 NFL Football"
author: "Harshavardhan Srijay"
date: "10/13/2020"
output:
  html_document: default
  pdf_document: default
---
```{r}
library(tidyverse)
```

# Import Data
```{r}
data <- read.csv(file = "nfl2020-week1.csv")
#Fix the actor in this play
data$actors[1586] = "Ryan Fitzpatrick"
```

I added a column to the original data on Excel called "EP_DIFF", which is the expected points after minus expected points before, for each play (EPA-EPB). I used this as an approximate measure of the expected points added by a given actor in a play.

```{r}
glimpse(data)
```

# Yards Per Play
```{r, fig.width = 10}
#summarise yards per play for each team
ypp <- data %>%
  filter(actions == "pass" | actions == "run" | actions == "sack") %>%
  mutate(dist = ifelse(is.na(distance), 0, distance)) %>%
  group_by(possessionStart) %>%
  summarise(mean = mean(dist))

#plot
ggplot(data = ypp, mapping = aes(x = reorder(possessionStart, mean), y = mean)) +
  geom_point(size = 3) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Yards Per Play for Each NFL Team", y = "Mean Yards Per Play", subtitle = "Week 1 - 2020 NFL Season") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank())
```

# Identify Positions

### Identify QBs
```{r}
qbs <- c()
for(i in 1:2788){
  if(data$actions[i] == "pass"){
    qbs <- c(qbs, data$actors[i])
  }
}
qbs <- unique(qbs)
```
### mutate position columns
```{r}
data <- data %>%
  mutate(qb = ifelse(actors %in% qbs, "QB", "non-QB")) %>%
  mutate(rb = ifelse(!(actors %in% qbs) & (actions == "run"), "RB", "non-RB" )) %>%
  mutate(wr = ifelse(is.na(receiver) | receiver == "UNKNOWN", "non-WR", "WR"))

#make columns factors
data$qb = as.factor(data$qb)
data$rb = as.factor(data$rb)
data$wr = as.factor(data$wr)
```

## Expected Points Added by Quarterbacks
```{r, fig.width = 10}
qbs <- data[data$qb == "QB",]
qbs$actors = as.factor(qbs$actors)
qbs$EP_DIFF = as.numeric(qbs$EP_DIFF)

#filter out qbs with less than 7 plays, to remove noise
high_freq <-  as.data.frame(table(qbs$actors)) %>%
  filter(Freq >= 7)
#summarise qb epa
top_qbs <- qbs %>%
  filter(qbs$actors %in% high_freq$Var1) %>%
  group_by(actors) %>%
  summarise(epa = mean(EP_DIFF), var = sd(EP_DIFF)) %>%
  arrange(desc(epa)) 
#plot
ggplot(data = top_qbs, mapping = aes(x = reorder(actors, epa), y = epa)) +
  geom_point(aes(size = var)) +
  coord_flip() +
  theme_minimal() +
  labs(y = "Mean Expected Points Added Per Posession", title = "Mean Expected Points Added per Play by Quarterbacks", subtitle = "Week 1 - 2020 NFL Season", size = "Standard Deviation") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank())
```

## Expected Points Added by Runningbacks
```{r, fig.width = 10}
rbs <- data[data$rb == "RB",]
rbs$actors = as.factor(rbs$actors)
rbs$EP_DIFF = as.numeric(rbs$EP_DIFF)

#filter out rbs with less than 7 plays, to remove noise
high_freq <-  as.data.frame(table(rbs$actors)) %>%
  filter(Freq >= 7)
#summarise rb epa
top_rbs <- rbs %>%
  filter(rbs$actors %in% high_freq$Var1) %>%
  group_by(actors) %>%
  summarise(epa = mean(EP_DIFF), var = sd(EP_DIFF)) %>%
  arrange(desc(epa)) %>%
  slice(1:30)
#plot
ggplot(data = top_rbs, mapping = aes(x = reorder(actors, epa), y = epa)) +
  geom_point(aes(size = var)) +
  coord_flip() +
  theme_minimal() +
  labs(y = "Mean Expected Points Added Per Posession", title = "Mean Expected Points Added per Play by Runningbacks", subtitle = "Week 1 - 2020 NFL Season", size = "Standard Deviation") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank())
```

## Expected Points Added by Receivers
```{r, fig.width = 10}
wr <- data[data$wr == "WR",]
wr$receiver = as.factor(wr$receiver)
wr$EP_DIFF = as.numeric(wr$EP_DIFF)

#filter out receivers with less than 7 plays, to remove noise
high_freq <-  as.data.frame(table(wr$receiver)) %>%
  filter(Freq >= 7)
#summarise receiver epa
top_wr <- wr %>%
  filter(receiver %in% high_freq$Var1) %>%
  group_by(receiver) %>%
  summarise(epa = mean(EP_DIFF), var = sd(EP_DIFF)) %>%
  arrange(desc(epa)) %>%
  slice(1:30)
#plot
ggplot(data = top_wr, mapping = aes(x = reorder(receiver, epa), y = epa)) +
  geom_point(aes(size = var)) +
  coord_flip() +
  theme_minimal() +
  labs(y = "Mean Expected Points Added Per Posession", title = "Mean Expected Points Added per Play by Receivers (WR and TE)", subtitle = "Week 1 - 2020 NFL Season", size = "Standard Deviation") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank())
```

## Third Down Conversion Rate
```{r, fig.width = 10}
data$possessionStart = as.factor(data$possessionStart)
third_down <- data %>%
  select(possessionStart, down, togo) 
third_down <- third_down[complete.cases(third_down),]

#identify all plays at third down, and plays at third down that had successful conversions
conversion <- array(0,dim=c(2411,1))
attempt <- array(0,dim=c(2411,1))

for (i in 1:2411){
  if(third_down$down[i] == 3 && third_down$possessionStart[i+1] == third_down$possessionStart[i]){
    attempt[i] = 1
  }
  if(third_down$down[i] == 3 && third_down$down[i+1] == 1 && third_down$possessionStart[i] == third_down$possessionStart[i+1]){
    conversion[i] = 1
  }
}
#summarise third down stats
third_down <- third_down %>%
  mutate(attempt = attempt) %>%
  mutate(conversion = conversion) %>%
  group_by(possessionStart) %>%
  summarise(total_conversion = sum(conversion), total_attempts = sum(attempt)) %>%
  mutate(conversion_rate = total_conversion/total_attempts)
#plot
ggplot(data = third_down, mapping = aes(x = reorder(possessionStart, conversion_rate), y = conversion_rate)) +
  geom_point(aes(size = total_attempts)) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Third Down Conversion Rate for NFL Teams", y = "Third Down Conversion Rate", subtitle = "Week 1 - 2020 NFL Season", size = "Total Attempts") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank()) +
  scale_y_continuous(labels = scales::percent)

```

## Field Goal Conversion Rate
```{r, fig.width = 10, warning=F}
data$actions = as.factor(data$actions)
#filter out non-FG plays
field_goal <- data %>%
  filter(actions == "FG") %>%
  select(possessionStart, actors, success, distance)
#summarise field goal attemps and conversions
field_goal$success = as.factor(field_goal$success)
field_goal <- field_goal %>%
  mutate(succ = ifelse(success == TRUE, 1, 0))
field_goal <- field_goal %>%
  group_by(possessionStart) %>%
  summarise(attempts = n(), success = sum(succ), dist = mean(distance)) %>%
  mutate(rate = success/attempts)
#plot
ggplot(data = field_goal, mapping = aes(x = reorder(possessionStart, rate), y = rate)) +
  geom_bar(stat = "identity", aes(fill = dist)) +
  scale_fill_gradient(low = "black",high = "dodgerblue")+
  coord_flip() +
  theme_minimal() +
  labs(title = "Field Goal Conversion Rate for NFL Teams", y = "Field Goal Conversion Rate", subtitle = "Week 1 - 2020 NFL Season", fill = "Average FG Distance (yd)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank()) +
  scale_y_continuous(labels = scales::percent)
```

## Passer Rating
```{r, fig.width = 10}
qbs$down <- as.factor(qbs$down)
#gather statistics to calculate passer rating
qbr <- qbs %>%
  filter(actions == "pass") %>%
  select(actors, success, interception, distance, score) %>%
  group_by(actors) %>%
  mutate(succ = ifelse(success == FALSE, 0, 1)) %>%
  mutate(int = ifelse(interception == FALSE | is.na(interception), 0, 1)) %>%
  mutate(dist = ifelse(is.na(distance), 0, distance)) %>%
  mutate(touch = ifelse(score == 6, 1, 0)) %>%
  summarise(attempts = n(), completions = sum(succ), passing_yards = sum(dist), interceptions = sum(int), touchdowns = sum(touch)) %>%
  filter(attempts > 10)
#calculate passer rating
qbr <- qbr %>%
  mutate(a = ((completions/attempts) - 0.3) * 5) %>%
  mutate(b = ((passing_yards/attempts) - 3) * 0.25) %>%
  mutate(c = (touchdowns/attempts) * 20) %>%
  mutate(d = 2.375 - ((interceptions/attempts) * 25))
qbr <- qbr %>% mutate(passer_rating = 100*((a+b+c+d)/6))
#plot
ggplot(data = qbr, mapping = aes(x = reorder(actors, passer_rating), y = passer_rating)) +
  geom_point(size = 3) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Passer Rating for NFL Quarterbacks", y = "Passer Rating", subtitle = "Week 1 - 2020 NFL Season") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), plot.subtitle = element_text(hjust = 0.5),axis.title.y = element_blank())
```

Clearly, some of the results displayed here are outcomes of noise, as all the data is only from the first week of NFL games. For instance, Gardner Minshew had a passer rating of almost 160, and Derrick Henry's expected points added was negative. Additionally, there was not much variation in the field goal conversion rates, as many teams had 100% conversion rates, since the sample sizes were so small. However, with the addition of play data from more games, the effect of random noise on the observed patterns should decrease, and the results should be more realistic.

## References
I obtained the formula for passer rating from Wikipedia: https://en.wikipedia.org/wiki/Passer_rating